An Extensible Approach to Verification of Embedded Systems

### 2014-05-01 tech-project/related_works_and_conclusion.tex
changes since initial turn in

There are a number of existing systems that ultimately seek to prove the functional correctness of programs. While we discuss only a handful here, readers interested in learning more about the systems mentioned might to refer to \cite{klebanov:2011} for a more in depth discussion. Perhaps most notable among the systems we list here (and perhaps the closest to RESOLVE's approach) is Microsoft research's Dafny \cite{leino:2010}. Designed to be capable of verifying pointer based structures, Dafny currently stands as one of the more popular approaches. Replacing RESOLVE's notion of mathematical variables with what are termed `ghost' variables, reasoning is kept, like RESOLVE, disconnected from implementation-level code. Dafny uses an intermediate verification language, termed Boogie, to generate verification conditions which are then dispatched via Z3, Microsoft's automated prover.

Another system worth mentioning here is VCC, also a Microsoft research project, which is designed for the verification of C programs.

@@

Using embedded application development as an example, we envision the work presented in this paper as a step towards more fully exploring the role of formal verification in lower level system-oriented contexts. Indeed, the realm of embedded systems has served as an ideal stepping off point for this challenge, as it encourages (or, indeed, requires) RESOLVE as a language to expand to provide support for not only new target languages (namely C), but also new language constructs that give the user freedom to more easily author drivers of the sort required for any serious embedded application. While we have made some progress with the latter in the form of the externally keyword, a natural direction for future work should involve moving beyond functional correctness, and instead focus more heavily on RESOLVE's mechanisms for specifying time and duration. With performance capabilities such as these in place, a natural next step in this work would be to formally specify and externally realize a simple memory allocation scheme, not dissimilar to the one presented in this paper.

Another language characteristic that we feel would benefit especially specification of lower level applications is having some means within RESOLVE of dictating \textit{static objects}. While currently RESOLVE allows users create single instance concepts that are tied to particular facility instantiations, we currently have no means of limiting how many instantiations of a particular facility there might be (not even via standard facilities). Thus, in terms of specifying concepts such as clocks, LED strips, and other hardware-centric concepts that only appear once on circuit board, RESOLVE might benefit from an in-language mechanism that restricts allowable instances of these and other objects.

\section{Conclusions}

In this paper we present new RESOLVE language constructs and tools with the objective of translating RESOLVE to embedded platform friendly C. We provide representative specifications of an LED strip component that both demonstrates the benefits of RESOLVE's approach to reusable mathematical developments, as well as formal verification of a single driver enhancement. We illustrate usage of this component in the context of a client facility module, wherein we introduce the externally keyword that allows users to transparently pair non-native RESOLVE implementations with native RESOLVE specifications. We then translate this facility application, illustrating a proof-of-concept dynamic memory allocator working in tandem with what we consider to be an accurate C representation of RESOLVE code. The examples and results listed in this paper have been successfully run on a Telos mote, and the overall tool -- now fully integrated with the RESOLVE compiler -- actively continues to undergo improvements and optimizations on all fronts.

>>>>

There are a number of existing systems that ultimately seek to prove the functional correctness of programs. While we discuss only two here, readers interested in learning more about the systems mentioned might to refer to \cite{klebanov:2011} for a more in depth discussion. Perhaps most notable among the systems we list here (and perhaps the closest to RESOLVE's approach) is Microsoft research's Dafny \cite{leino:2010}. Designed to be capable of verifying pointer based structures, Dafny currently stands as one of the more popular approaches to verification. Substituting RESOLVE's notion of mathematical variables with what are termed `ghost' variables, reasoning is kept, like RESOLVE, disconnected from implementation-level code. Dafny uses an intermediate verification language, termed Boogie, to generate verification conditions which are then dispatched via Z3, Microsoft's automated prover.

Another system, also a Microsoft research endeavor, is VCC -- a system designed for the verification of C programs \cite{vcc:2009}. One of the main draws of VCC is that it has mechanisms for verification of concurrent code -- an ideal feature when presented with asynchronous, event driven code common in most embedded applications.

RESOLVE differs from the projects mentioned above most notably in that it does not make use of third party automated provers or decision procedures to discharge verification conditions. Rather, it makes use of an integrated, minimalist term rewrite prover that applies higher order theorems and definitions in attempting to transform givens into a provable goal.

@@

Using embedded application development as an example, we perceive this work as a step towards more fully exploring the role of formal verification in lower level system-oriented contexts. Indeed, the realm of embedded systems provides an ideal stepping off point for this challenge, as it encourages (and indeed, requires) RESOLVE to provide support for not only new target languages (namely C), but also new language constructs that give the user the ability to author drivers required for any serious embedded application. While we have made progress with the latter in the form of the externally keyword, a natural direction for future work would entail using RESOLVE's performance specification mechanisms to prove the time and duration estimates for a memory allocator, not unlike the one presented here.

Further, an additional language characteristic that we feel would be beneficial is having some means of dictating \textit{static objects}. While RESOLVE makes considerations for single instance concepts that are tied to particular facility instantiations, we currently have no means of strictly limiting how many instantiations of a particular facility there might be (not even via standard facilities). Thus, in terms of specifying concepts such as clocks, LED strips, and other hardware-centric concepts that only appear once on a circuit board, RESOLVE could benefit from a mechanism that restricts allowable instances of these and other objects.

\section{Conclusion}

In this paper we present a new RESOLVE language construct, and a translation tool that aims to produce an accurate, embedded platform-friendly C representation of RESOLVE. We provide representative specifications of an LED strip component that both demonstrates the benefits of RESOLVE's approach to reusable mathematical developments, as well as formal verification of a single driver enhancement. We illustrate usage of this component in the context of a client facility module, wherein we introduce the externally keyword that allows users to transparently pair non-native RESOLVE implementations with native RESOLVE specifications. We then translate this facility application, illustrating a proof-of-concept dynamic memory allocator working in tandem with what we consider to be an accurate C representation of RESOLVE code. The examples and results listed in this paper have been successfully run on a Telosb mote, and the overall tool -- now fully integrated with the RESOLVE compiler -- actively continues to undergo improvements and optimizations on all fronts.

### 2014-05-01 tech-project/specify_led.tex
changes since initial turn in

\subsection{Concepts}
\label{ssec:concepts}
In RESOLVE, programs are composed of several different modules that range from interfaces and realizations, to client (facility) modules. A \textit{concept} module in RESOLVE defines a specification for a mathematical, abstract type. Similar to an interface in Java, a concept provides a number of operation signatures that implementors are expected to realize. Shown below is a \texttt{LED\_Template} concept that provides a light strip abstraction.

\begin{verbatim}
Concept LED_Template(eval Strip_Length: Integer);
    uses Boolean_Theory, String_Theory;
    requires 0 < Strip_Length <= 4;

@@

    Oper Set(updates L : LED; eval b : Boolean; 
                             eval i : Integer);
        requires 0 <= i < |L|;
        ensures |L| = |#L| and 
                   Element_At(i, L) = b;

    Oper Status(preserves L : LED; 
                   eval i : Integer) : Boolean;
        requires 0 <= i < |L|;
        ensures Status = Element_At(i, L);

@@

%The \texttt{Type Family} clause introduces an abstract type for the LED strip which we mathematically model using strings of booleans, denoted \texttt{Str(B)}, while the \texttt{exemplar} declared immediately after provides a handle to this type family. Use of the \texttt{exemplar} can be observed in the \texttt{constraint} clause, where it is used to assert that length of the strip is fixed to be whatever the user has specified via the \texttt{Strip\_Length} parameter. Finally, the \texttt{initialization} \texttt{ensures} clause makes explicit the state of the structure at the time it is initialized. We omit this verification specific clause, as it has little bearing on verification due to the fact that the realization we provide is \textit{externally} realized -- a point discussed in the next section\footnote{This clause, if it were provided, would need to communicate the following: ``For each position $i$ within the strip, the LED (or, value) at position $i$ is initially false."}

It's worth noting that unlike the \texttt{LED\_Template} presented in \cite{regula:2010} which models an LED as the \textit{cartesian product} of booleans $b_0$, $b_1$, \ldots , $b_4$, the strip model we present here instead uses strings for the following reasons:
\begin{itemize}
\item Strings are indexible, and thus do not require separate \texttt{Set} and \texttt{Status} operations for each individual LED.
\item This approach demonstrates the benefits of reusable mathematical theories. The specifications listed here are based (almost) entirely in \texttt{String\_Theory} and are therefore able to make use of RESOLVE's pre-existing math libraries.
\end{itemize}

Finally, the concept provides two operations. The first, \texttt{Set}, takes as a parameter an instance of an LED strip \texttt{L}, a boolean \texttt{b}, and an integer \texttt{i}. The operation \texttt{requires} that \texttt{i} falls within the length of the strip, and \texttt{ensures} two things upon completion: The length of the outgoing strip \texttt{L} is the same as the incoming strip, \texttt{\#L}, and that the LED in position \texttt{i} of \texttt{L} is set to boolean \texttt{b}. The \texttt{Status} operation is specified similarly.

%Each operation carries with it a pre and post condition in the form of \texttt{requires} and \texttt{ensures} clause, which make explicit what must be true coming into the function, and what must be true after. For instance, the requires clause to \texttt{Status} states that \texttt{i} must fall somewhere within the bounds of the string which models our LED strip, while the \texttt{ensures} clause makes use of a \texttt{String\_Theory} definition, \texttt{Element\_At}, to assert that the boolean returned from \texttt{Status} is indeed the value occupying position \texttt{i} of \texttt{L}. It's important to note that the variables appearing in the \texttt{requires} and \texttt{ensures} clauses are strictly mathematical values. Thus, the \texttt{L} parameter of operation \texttt{Status} is referring to the mathematical value of an \texttt{LED}, as opposed to a programmatic one (typically defined in a realization).

@@

\begin{verbatim}
Enhancement Toggling_Capability for LEDs_Template;

    Oper Toggle(upd L : LED; eval i : Integer);
        requires 0 <= i < |L|;
        ensures Element_At(i, L) = 
                        not(Element_At(i, #L));
end Toggling_Capability;

Realization Toggling_Realiz for
            Toggling_Capability of LEDs_Template;

    Proc Toggle(upd L : LED; eval i : Integer);
        Var Content : Boolean;

        Content := Status(L, Replica(i));
        Set(L, Not(Content), Replica(i));
    end Toggle;

@@

Enhancement realizations are neutral as well since any method called within the context of an enhancement realization refers to the operation specified in the concept -- meaning no knowledge of implementation details is required.

@@

We now turn to the task of verifying our small toggling enhancement. The first step in doing so is to generate Verification Conditions (VCs) for \texttt{Toggle\_Realiz}, which, if proven, will establish the correctness of this particular implementation. One thing to note about the VCs themselves is that they are generated from specific lines of a realization, and exist to ensure that the content of the realization is consistent with its specification: This entails checking for things such as array access boundary violations, etc.

\begin{figure}[!htb]
\centering
\begin{tabular}{lccc}
	\toprule
	Condition \# & Time (ms)	& Steps	& Search \\
	\midrule
	VC 0\_1	& 4426	& 5	& 0	\\
	VC 0\_2	& 5039	& 5	& 0	\\
	VC 0\_3	& 6324	& 6	& 0	\\
	\bottomrule
\end{tabular}
\caption{Verification results for operation \texttt{Toggle}}
\label{fig:results}
\end{figure}

As the results summarized in Figure \ref{fig:results} indicate, using RESOLVE's integrated prover, we are able to mechanically and automatically dispatch all VCs for \texttt{Toggling\_Realiz}, thus verifying its correctness. In terms of proof difficulty, given the number of steps and time taken to establish each, we conclude that the VCs generated were of a straightforward variety. Readers interested however in learning more about the steps and specific actions the prover takes in transforming givens and dispatching similar (and more complex) VCs should refer to \cite{smith:2013}.

@@

        (* Declare LED strip indices *)
        Var I1, I2, I3 : Integer;
        Var Loop : Boolean;

@@

        I1 := 1; I2 := 2; I3 := 3;

        Loop := True();
        While(Loop)
            changing Loop;
            maintaining ...
        do
            Leds_Fac.Toggle(L, I1);
            Std_Clock_Fac.Wait_500_Milli_Seconds();

            Leds_Fac.Toggle(L, I2);
            Std_Clock_Fac.Wait_500_Milli_Seconds();

            Leds_Fac.Toggle(L, I3);
            Std_Clock_Fac.Wait_500_Milli_Seconds();
        end;
    end Main;

end LED_Telos_Demo;
\end{verbatim}
%smaller file sizes
%reusable code
%only have to translate a file once
% be able to say that the code we have is the code that was verified.
% our approach takes less ROM. Its not less efficient, it is differently efficient.
Prior to using the LED component developed in the previous sections, we first must bind our \texttt{LED\_Template} specification with an appropriate realization. This is accomplished via the facility declaration located directly beneath the \texttt{uses} clause, which pairs the specification (\texttt{LED\_Template}) with a realization (\texttt{Std\_Led\_Realiz}). Note that the ability to toggle is added \textit{on top} of this facility declaration in a similar fashion. The remaining bulk of logic driving the application rests in the non terminating busy loop inside operation \texttt{Main}, where we use our enhancement-provided \texttt{Toggle} operation to successively turn each light within the strip on, then off.

@@

\begin{figure*}[!htb]
\centering
\includegraphics[scale=.55]{figs/component_graph.pdf}
\caption{A diagram illustrating inter-component relationships for the example in Section \ref{sec:specifying}}
\end{figure*}
\label{fig:imp}

Readers might note that we never presented a realization of \texttt{LED\_Template}. Indeed, after having written the concept, the RESOLVE programmer would ideally provide it with a verifiable, native RESOLVE implementation. However, as our target platform is embedded, and our concept aims to provide control for LEDs -- a decidedly low level feature on embedded hardware -- our realization is forced to operate at similarly low levels by directly manipulating hardware pins provided by the msp430's chipset.

@@

We feel this new keyword is beneficial for the following reasons:
\begin{itemize}
\item The language no longer must ``hide" the fact that some of the lower level components relied upon are not written in straight-line, native RESOLVE code. The ``externally" keyword now transparently indicates this.
\item Provides flexibility for those users looking to wrap their low-level programs/drivers with formal RESOLVE interface specifications.
\end{itemize}

>>>>

\subsection{Concepts}
\label{ssec:concepts}
In RESOLVE, programs are composed of several different modules that range from interfaces and realizations, to client (facility) modules. The first we  discuss here, termed a \textit{concept}, defines a specification for a mathematical, abstract type. Similar to an interface in Java, a concept provides a number of operation signatures that implementors are expected to realize. Shown below is an \texttt{LED\_Template} concept that provides an LED light strip abstraction.

\begin{verbatim}
Concept LED_Template(eval Strip_Length: Integer);
    uses Boolean_Theory, String_Theory;
    requires 0 < Strip_Length

@@

    Oper Set(updates L : LED; eval b : Boolean; 
                             eval i : Integer);
        requires 0 <= i < |L|;
        ensures L = (Prt_Btwn(0, i, #L) o <#b>) o
            Prt_Btwn(i, |#L|, #L) and |L| = |#L|;

    Oper Status(preserves L : LED; 
                        updates b : Boolean;
            eval i : Integer);
        requires 0 <= i < |L|;
        ensures Status = Element_At(i, #L);

@@

It's worth noting that unlike the \texttt{LED\_Template} presented in \cite{regula:2010} which models an LED as the \textit{cartesian product} of booleans $b_0$, $b_1$, \ldots , $b_4$, the strip model we present here instead uses strings for the following reasons:
\begin{itemize}
\item Strings are indexible, and thus do not require separate \texttt{Set} and \texttt{Status} operations for each individual LED.
\item This approach demonstrates the benefits of reusable mathematical theories. Indeed, the specifications listed here are based (almost) entirely in \texttt{String\_Theory} and are therefore able to make use of RESOLVE's pre-existing math libraries.
\end{itemize}

The concept provides two operations. The first, \texttt{Set}, takes as a parameter an instance of an LED strip \texttt{L}, a boolean \texttt{b}, and an integer \texttt{i}. The operation \texttt{requires} that \texttt{i} falls within the length of the strip, and \texttt{ensures} two things upon completion: The length of the outgoing strip \texttt{L} is the same as the incoming strip, \texttt{\#L}, and that the LED in position \texttt{i} of \texttt{L} is set to boolean \texttt{b}\footnote{Note that the \texttt{ensures} clause in this case is strong enough to assert that \textit{only} the boolean at location \texttt{i} has changed}. The \texttt{Status} operation is specified similarly.

@@

\begin{verbatim}
Enhancement Toggling_Capability for LED_Template;

    Oper Toggle(upd L : LED; eval i : Integer);
        requires i >= 0 and i < |L|
        ensures L = (Prt_Btwn(0, i, #L) o  
                  <not(Element_At(i, #L))>) o
                  Prt_Btwn(i, |#L|, #L);
end Toggling_Capability;

Realization Toggling_Realiz for
            Toggling_Capability of LED_Template;

    Proc Toggle(upd L : LED; eval i : Integer);
        Var Flipped_Contents : Boolean;
        Status(L, Flipped_Contents, Replica(i));
        Set(L, Not(Flipped_Contents), Replica(i));
    end Toggle;

@@

Enhancement realizations are similar in the sense that any operation called within the context of an enhancement realization refers to the operation specified in the concept -- meaning no knowledge of its implementation details are required.

@@

We now turn to the task of verifying our small toggling enhancement. The first step in doing so is to generate Verification Conditions (VCs) for \texttt{Toggle\_Realiz}, which, if proven, will establish the correctness of this particular implementation. One thing to note about the VCs themselves is that they are generated from specific lines of a realization, and exist to ensure that the content of the realization is consistent with its specification: This entails checking for array access violations, parameters not meeting preconditions, etc.

\begin{figure}[!htb]
\centering
\begin{tabular}{lccc}
	\toprule
	Condition \# & Time (ms)	& Steps	& Search \\
	\midrule
	VC 0\_1	& 7009	& 6	& 0	\\
	VC 0\_2	& 5185	& 5	& 0	\\
	VC 0\_3	& 3354	& 6	& 0	\\
	VC 0\_4	& 4810	& 5	& 0	\\
	VC 0\_5	& 7133	& 5	& 0	\\
	\bottomrule
\end{tabular}
\caption{Verification results for operation \texttt{Toggle}}
\label{fig:results}
\end{figure}

As the results summarized in Figure \ref{fig:results} indicate, using RESOLVE's integrated prover, we are able to mechanically and automatically dispatch all VCs for \texttt{Toggling\_Realiz}, thus verifying its correctness. In terms of proof difficulty, given the number of steps and time taken to establish each goal, we conclude that the VCs generated are of a straightforward variety. Readers interested however in learning more about the steps and specific actions the prover takes in transforming givens and dispatching similar (and more complex) VCs should refer to \cite{smith:2013}.

@@

        (* Declare LED strip indices *)
        Var I0, I1, I2 : Integer;
        Var Loop : Boolean;

@@

        I0 := 0; I1 := 1; I2 := 2;

        Loop := True();
        While(Loop)
            changing Loop;
            maintaining ...
        do
            Leds_Fac.Toggle(L, I0);
            Std_Clock_Fac.Wait_500_Milli_Seconds();

            Leds_Fac.Toggle(L, I1);
            Std_Clock_Fac.Wait_500_Milli_Seconds();

            Leds_Fac.Toggle(L, I2);
            Std_Clock_Fac.Wait_500_Milli_Seconds();
        end;
    end Main;

end LED_Telos_Demo;
\end{verbatim}

Prior to using the LED component developed in the previous sections, we first must bind our \texttt{LED\_Template} specification with an appropriate realization. This is accomplished via the facility declaration located directly beneath the \texttt{uses} clause, which pairs the specification (\texttt{LED\_Template}) with a realization (\texttt{Std\_Led\_Realiz}). Note that the ability to toggle is added \textit{on top} of this facility declaration in a similar fashion. The remaining bulk of logic driving the application rests in the non terminating busy loop inside operation \texttt{Main}, where we use our enhancement-provided \texttt{Toggle} operation to successively turn each LED within the strip on, then off.

@@

\begin{figure*}[!htb]
\centering
\includegraphics[scale=.55]{figs/component_graph.pdf}
\caption{A diagram illustrating inter-component relationships for the example detailed in Section 3. Files displayed in clouds indicate externally realized components, while dotted line separating these from others indicates the current boundary between low-level, hardcoded components, and native RESOLVE abstractions.}
\end{figure*}
\label{fig:imp}

Readers might note that we never presented a realization of \texttt{LED\_Template}. Indeed, after having written the concept, the RESOLVE programmer would ideally provide a verifiable, native RESOLVE implementation. However, as our target platform is embedded, and our concept aims to provide control for LEDs -- a decidedly low level feature on embedded hardware -- our realization is forced to operate at similarly low levels by directly manipulating hardware pins provided by the msp430's chipset.

@@

We feel this new keyword is beneficial for the following reasons:
\begin{itemize}
\item The language no longer must ``hide" the fact that some of the lower level components relied upon are not written in straight-line, native RESOLVE. The ``externally" keyword now transparently indicates this.
\item Provides flexibility for those users looking to wrap their low-level programs/drivers with formal RESOLVE interface specifications.
\end{itemize}
